### 2016-07-03 reasoning.tex
Typos fixed.

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away 
environmental interference as a rely relation. A conspicuous
difference between a standard development of rely-guarantee and ours
is that, while the former reasons in terms of program states (variable
to value bindings), we reason in terms of executions as captured by
their traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and pre- ($P$) and post- ($Q$) conditions as assertions over executions
(i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on states are also written
similarly. For instance, consider the postcondition
$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{C} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{C} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:

@@

The standard rely-guarantee judgment is a quintuple
$\rg{P,R}{c}{G,Q}$, which informally asserts that if a
command $c$ is executed in a state that satisfies its precondition
$P$, provided that every interference step during the execution is
contained inside the rely relation $R$, then the effect that each step
of executing $c$ has on the state is captured by $G$, and the final
state of execution satisfies the postcondition $Q$. The terms
``execution'' and ``step'' are defined with respect to an operational
semantics, which, in our case, is parameterized on the trace invariant
$\I$. Consequently, our rely-guarantee judgment for commands is the
sextuple $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the
quintuple in that it requires every step of the execution to additionally preserve
the trace invariant $\I$ for the postcondition to be valid.  As usual,
if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$. \txnimp
expressions are side-effecting (they generate \C{RD} effects), and
admit interference during their evaluation.  Therefore, like 
commands, expressions need to be reasoned about explicitly.  In contrast to
commands, however, expressions evaluate to values, and the reasoning
framework should also admit assertions on such values.  We therefore
define a separate judgment for expressions - the septuple
$\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
\Prop$ is an assertion on the return value of $e$.

@@

\noindent However, the assumption that interference preserves $\I$ (or,
dually, $\I$ withstands inteference) needs to be justified
separately. We call this the stability requirement on $\I$:

@@

\noindent The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
shared variable reads inside a transaction $T_i$. It requires $\I$ to
be stable, and pre- and post- conditions to be stable relative to
$\I$.  The quantified premise effectively requires a proof that if the
abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
an execution $\E$ that satisfies the precondition $P$, then the
resultant execution $\E'$ satisfies the postcondition $Q$, and that
the guarantee $G$ faithfully captures the transition from $\E$ to
$\E'$. It is informative to compare this premise with the premise of
the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
premises also appear in \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
rules, which define rely-guarantee judgments for assignments and
transactions, respectively. \rulelabel{RG-Var} however also requires
the return value ($\interp{S}(X)$) of the read to satisfy the
assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the RG
judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
corresponding judgments for the constitutent expressions $e_1$ and
$e_2$. The quantified premise requires any value resulting from
evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
$e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
$C_2$, respectively. The rules for sequential and parallel composition
of commands are essentially the same as their counterparts in a
standard rely-guarantee formulation and hence elided.

@@

The \rulelabel{RG-Conseq} rule define ways to strengthen or
weaken relations and assertions associated with the RG judgment of
transaction-bound expressions. Similar rules exist for
transaction-bound and top-level commands, but are not
shown.\footnote{The supplementary materials provide the complete set
  of rules.} As is the case with a standard rely-guarantee
formulation,the rules allow the precondition $P$ and the rely relation
$R$ to be strengthened, and the postcondition $Q$ (also, $C$ in the
case of expressions) and the guarantee relation $G$ to be weakened.
The most notable aspect of the \rulelabel{RG-Conseq} rules is that
they allow the trace invariant $\I$ to be strengthened. Considering
that $\I$ captures isolation properties, this means that a program
proven correct under weaker isolation levels is also correct under
stronger ones.  Parametricity over the trace invariant $\I$, combined
with the ability to strengthen $\I$ as needed, allows our proof system
to support a highly flexible proof strategy to prove programs correct
over various isolation variants. For example, programmers can
\emph{define} isolation guarantees of their choice (by defining $\I$
appropriately) and then prove programs correct assuming the guarantees
hold.  The soundness of strengthening $\I$ ensures that a program can
be executed safely on any data store that offers isolation guarantees
at least as strong as those assumed.

@@

Note that the semantics of all RG judgments, including the judgments
for transaction-bound terms, make similar demands of the guarantee
relation. Given that transactions are atomic (though not isolated), it
is not immediately apparent why a transaction's guarantee is required
to make explicit every step of its reduction. This requirement is
justifed however because, in reality, a transaction's atomicity is
predicated on the isolation settings of the observer. A \iso{Read
  Uncommitted} transaction, for example, is permitted to observe the
internal state of a transaction $T$ even if $T$ is claimed to execute
atomically.  In the interest of modular verification, the transaction
must therefore make its internal state available via its guarantee
relation.

>>>>

As mentioned earlier, our proof system is a rely-guarantee framework
that admits compositional reasoning by abstracting away 
environmental interference as a rely relation. A conspicuous
difference between a standard development of rely-guarantee and ours
is that, while the former reasons in terms of program states (variable
to value bindings), we reason in terms of executions as captured by
their traces ($\E$). In particular, our rely ($R$) and guarantee ($G$)
relations relate executions (i.e., $R,\,Q\,\subseteq\,\E\times\E$),
and pre- ($P$) and post- ($Q$) conditions as assertions over executions
(i.e., $P,\,Q\, : \E \rightarrow \Prop$). Our development nonetheless
facilitates state-based reasoning via the interpretation function
($\interp{\cdot}$) introduced in \S\ref{sec:syntax}. For example, if a
\emph{bi-state} rely relation relates every pair of states $\sigma$
and $\sigma'$ such that $\sigma'(X) \ge \sigma(X) \ge 0$, the
corresponding \emph{bi-execution} rely relation relates every pair of
executions $\E$ and $\E'$ such that $\interp{\E}(X) \ge
\interp{\E'}(X) \ge 0$.  Assertions on states are also written
similarly. For instance, consider the post-condition
$\C{C\,=\,k-a1-a2}$ of the program in Fig.~\ref{fig:motiv-eg-1}. The
corresponding assertion on the post-state
($\lambda\sigma.~\sigma(\C{C}) \,=\, \C{k-a1-a2}$) asserts that in all
states resulting from executing the program, the value of \C{C} is
\C{k-a1-a2}. The equivalent execution-based assertion
($\lambda\E.~\interp{\E}(\C{C})\,=\,\C{k-a1-a2}$) asserts that in all
executions of the program, the value of \C{C} is \C{k-a1-a2}.
However, having access to an execution facilitates assertions that go
beyond the state. An example is an invariant of \C{Wd1} described
abstractly in \S\ref{sec:motivation}, and reified as an
execution-based assertion below:

@@

The standard rely-guarantee judgment is a quintuple
$\rg{P,R}{c}{G,Q}$, which informally asserts that if a
command $c$ is executed in a state that satisfies its precondition
$P$, provided that every interference step during the execution is
contained inside the rely relation $R$, then the effect that each step
of executing $c$ has on the state is captured by $G$, and the final
state of execution satisfies the post-condition $Q$. The terms
``execution'' and ``step'' are defined with respect to an operational
semantics, which, in our case, is parameterized on the trace invariant
$\I$. Consequently, our rely-guarantee judgment for commands is the
sextuple $\rg{\I,P,R}{c}{G,Q}$, whose semantics differs from the
quintuple in that it requires every step of the execution to additionally preserve
the trace invariant $\I$ for the post-condition to be valid.  As usual,
if $c$ is inside a transaction $T_i$, we write $\txnbox{c}_i$. \txnimp
expressions are side-effecting (they generate \C{RD} effects), and
admit interference during their evaluation.  Therefore, like 
commands, expressions need to be reasoned about explicitly.  In contrast to
commands, however, expressions evaluate to values, and the reasoning
framework should also admit assertions on such values.  We therefore
define a separate judgment for expressions - the septuple
$\rg{\I,P,R}{e}{G,C,Q}$. The new entrant $C : \mathbb{N} \rightarrow
\Prop$ is an assertion on the return value of $e$.

@@

\noindent However, the assumption that interference preserves $\I$ (or,
dually, $\I$ withstands interference) needs to be justified
separately. We call this the stability requirement on $\I$:

@@

\noindent The rule \rulelabel{RG-Var} defines the rely-guarantee judgment for
shared variable reads inside a transaction $T_i$. It requires $\I$ to
be stable, and pre- and post- conditions to be stable relative to
$\I$.  The quantified premise effectively requires a proof that if the
abstract machine of Fig.~\ref{fig:txnimp} takes a step starting from
an execution $\E$ that satisfies the precondition $P$, then the
resultant execution $\E'$ satisfies the post-condition $Q$, and that
the guarantee $G$ faithfully captures the transition from $\E$ to
$\E'$. It is informative to compare this premise with the premise of
the \rulelabel{E-Aux} reduction rule of Fig.~\ref{fig:txnimp}. Similar
premises also appear in \rulelabel{RG-Asgn} and \rulelabel{RG-Txn}
rules, which define rely-guarantee judgments for assignments and
transactions, respectively. \rulelabel{RG-Var} however also requires
the return value ($\interp{S}(X)$) of the read to satisfy the
assertion $C$ meant for the value. \rulelabel{RG-Arith} defines the RG
judgment for an arithmetic expression $e_1\pm e_2$ in terms of the
corresponding judgments for the constituent expressions $e_1$ and
$e_2$. The quantified premise requires any value resulting from
evaluating $e_1 \pm e_2$ to satisfy the assertion $C$, provided that
$e_1$ and $e_2$ always evaluate to values that satisfy $C_1$ and
$C_2$, respectively. The rules for sequential and parallel composition
of commands are essentially the same as their counterparts in a
standard rely-guarantee formulation and hence elided.

@@

The \rulelabel{RG-Conseq} rule define ways to strengthen or
weaken relations and assertions associated with the RG judgment of
transaction-bound expressions. Similar rules exist for
transaction-bound and top-level commands, but are not
shown.\footnote{The supplementary materials provide the complete set
  of rules.} As is the case with a standard rely-guarantee
formulation,the rules allow the precondition $P$ and the rely relation
$R$ to be strengthened, and the post-condition $Q$ (also, $C$ in the
case of expressions) and the guarantee relation $G$ to be weakened.
The most notable aspect of the \rulelabel{RG-Conseq} rules is that
they allow the trace invariant $\I$ to be strengthened. Considering
that $\I$ captures isolation properties, this means that a program
proven correct under weaker isolation levels is also correct under
stronger ones.  Parametricity over the trace invariant $\I$, combined
with the ability to strengthen $\I$ as needed, allows our proof system
to support a highly flexible proof strategy to prove programs correct
over various isolation variants. For example, programmers can
\emph{define} isolation guarantees of their choice (by defining $\I$
appropriately) and then prove programs correct assuming the guarantees
hold.  The soundness of strengthening $\I$ ensures that a program can
be executed safely on any data store that offers isolation guarantees
at least as strong as those assumed.

@@

Note that the semantics of all RG judgments, including the judgments
for transaction-bound terms, make similar demands of the guarantee
relation. Given that transactions are atomic (though not isolated), it
is not immediately apparent why a transaction's guarantee is required
to make explicit every step of its reduction. This requirement is
justified however because, in reality, a transaction's atomicity is
predicated on the isolation settings of the observer. A \iso{Read
  Uncommitted} transaction, for example, is permitted to observe the
internal state of a transaction $T$ even if $T$ is claimed to execute
atomically.  In the interest of modular verification, the transaction
must therefore make its internal state available via its guarantee
relation.

### 2016-07-03 related-work.tex
Case studies.
Bib fixed. Case studies edited.
Typos fixed.

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specified several weak isolation levels in terms
of \emph{dependency graphs} between transactions, by defining the
kinds of dependency cycles that are forbidden in each case. The
operational nature of Adya's specifications make them suitable for
runtime monitoring and anamoly
detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
declarative nature of our specifications make them suitable for
reasoning about program behaviour. Sivaramakrishnan \emph{et
  al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, and is thus not
suitable to reason operationally about a program as it builds the
trace. Cerone \emph{et al.}~\cite{gotsmanconcur15} present a framework
to specify isolation levels with atomic visibility, but their
specifications are also for complete traces; Moreover, their
formulation implicitly assumes homogenity, i.e., they require all
transactions to execute using either {\sc si} or {\sc ser}, which is
often not the case in practice. While all the aforementioned
specification frameworks use the vocabulary introduced
in~\cite{burckhardt14} to specify replicated data types and different
forms of weak consistency, in contrast to this work, none of them come
with an associated reasoning framework that can use such
specifications productively towards verifying programs under weak
isolation.

@@

\paragraph{Reasoning under relaxed memory} Considering the
correspondences between axiomatic models of relaxed
memory~\cite{battycpp} and weakly consistent
replication~\cite{burckhardt14}, it is reasonable to expect similar
reasoning approaches to work in both the cases. Ridge~\cite{rgtso}
generalizes rely-guarantee reasoning to the x86-TSO memory model.
Likewise, Vafeadis \emph{et al.}~\cite{rsl13} generalize concurrent
separation logic (CSL)~\cite{csl} to the C11 relaxed memory model.
Ferreira \emph{et al.}~\cite{ferreira10} propose a parameterized
operational semantics for relaxed memory models, but the
parameterization is over a relation between relaxed memory programs
and related {\sc sc} programs. They also prove the soundness of CSL
for the weakest memory model expressible in their semantics, but
neither a parameterized (in the aforementioned sense) CSL, nor a CSL
specialized for any relaxed memory model is proposed.  ~\cite{DLZ+13}
presents a \emph{buffered memory model} for Java that defines an
axiomatic definition for the JMM in terms of memory reorderings, and
an operational instantiation consistent with the TSO memory model
found on x86 processors, and demonstrates the equivalence of the two
definitions.  However, there is no attempt to parameterize the model
over different reordering or consistency definitions, which is a central
focus of this paper.

>>>>

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specified several weak isolation levels in terms
of \emph{dependency graphs} between transactions, by defining the
kinds of dependency cycles that are forbidden in each case. The
operational nature of Adya's specifications make them suitable for
runtime monitoring and anomaly
detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
declarative nature of our specifications make them suitable for
reasoning about program behaviour. Sivaramakrishnan \emph{et
  al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, and is thus not
suitable to reason operationally about a program as it builds the
trace. Cerone \emph{et al.}~\cite{gotsmanconcur15} present a framework
to specify isolation levels with atomic visibility, but their
specifications are also for complete traces; Moreover, their
formulation implicitly assumes homogeneity, i.e., they require all
transactions to execute using either {\sc si} or {\sc ser}, which is
often not the case in practice. While all the aforementioned
specification frameworks use the vocabulary introduced
in~\cite{burckhardt14} to specify replicated data types and different
forms of weak consistency, in contrast to this work, none of them come
with an associated reasoning framework that can use such
specifications productively towards verifying programs under weak
isolation.

@@

\paragraph{Reasoning under relaxed memory} Considering the
correspondences between axiomatic models of relaxed
memory~\cite{battycpp} and weakly consistent
replication~\cite{burckhardt14}, it is reasonable to expect similar
reasoning approaches to work in both the cases. Ridge~\cite{rgtso}
generalizes rely-guarantee reasoning to the x86-TSO memory model.
Likewise, Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent
separation logic (CSL)~\cite{csl} to the C11 relaxed memory model.
Ferreira \emph{et al.}~\cite{ferreira10} propose a parameterized
operational semantics for relaxed memory models, but the
parameterization is over a relation between relaxed memory programs
and related {\sc sc} programs. They also prove the soundness of CSL
for the weakest memory model expressible in their semantics, but
neither a parameterized (in the aforementioned sense) CSL, nor a CSL
specialized for any relaxed memory model is proposed.  ~\cite{DLZ+13}
presents a \emph{buffered memory model} for Java that defines an
axiomatic definition for the JMM in terms of memory reorderings, and
an operational instantiation consistent with the TSO memory model
found on x86 processors, and demonstrates the equivalence of the two
definitions.  However, there is no attempt to parameterize the model
over different reordering or consistency definitions, which is a central
focus of this paper.
